## The math behind the chi-square goodness of fit test

```{block, type = "info"}
If the math below makes your eyes glaze over, you can skip it. This is presented for those who find it useful to understand the math behind the statistics to help understand what's happening.
```

We're going to continue using the card deck example. This data comes from the lsj-data dataset named "randomness". If you perform descriptive statistics of `choice_1` and ask for frequency tables, you get the *observed* frequencies. We expect the frequencies to be 25% for each choice (or *n* = 50 because our total N is 200, so 200/4 = 50).

|                    |               | $\clubsuit$ | $\diamondsuit$ | $\heartsuit$ | $\spadesuit$ |
|--------------------|---------------|-------------|----------------|--------------|--------------|
| Observed frequency | $O_i$         | 35          | 51             | 64           | 50           |
| Expected frequency | $E_i$         | 50          | 50             | 50           | 50           |
| Difference score   | $O_i-E_i$     | -15         | 1              | 14           | 0            |
| Difference squared | $(O_i-E_i)^2$ | 225         | 1              | 196          | 0            |

The formula for the chi-square goodness-of-fit test is as follows:

$\chi^2 = \sum{\frac{(O_i-E_i)^2}{E_i}} = \frac{225 + 1 + 196 + 0}{50} = \frac{422}{50} = 8.44$

Up to this point, we've been working with the *t* and *F* distributions for our statistics. Now we move to the $\chi^2$ distribution with $k-1$ degrees of freedom (k = number of groups; in our card's example case it's 4-1 = 3). What is the chi-square distribution? It looks a little something like this:

```{r}
library(ggplot2)
ggplot(data.frame(x = c(0, 30)), aes(x = x)) +
  stat_function(fun = dchisq, args = list(df = 3), aes(colour = "df = 03")) +
  stat_function(fun = dchisq, args = list(df = 4), aes(colour = "df = 04")) +
  stat_function(fun = dchisq, args = list(df = 5), aes(colour = "df = 05")) +
  stat_function(fun = dchisq, args = list(df = 10), aes(colour = "df = 10")) +
  stat_function(fun = dchisq, args = list(df = 20), aes(colour = "df = 20")) +
 # scale_color_viridis(discrete = TRUE) +
  theme_classic() +
  ylab(element_blank()) +
  xlab(label = "Value")
```

In our case, we have a degrees of freedom of 3. Just like we did early in the semester, we can look up our critical-$\chi^2$ value in a table to find that our critical value for df = 3 is 7.815. The figure below shows our critical-$\chi^2$ value in red and calculated-$\chi^2$ value in black. Because our calculated-$\chi^2$ value is in the critical region, we can reject the null hypothesis that all four suits are chosen with equal probability.

```{r}
ggplot(data.frame(x = c(0, 20)), aes(x = x)) +
  stat_function(fun = dchisq, args = list(df = 3)) +
  stat_function(fun = dchisq, args = list(df = 3), 
                xlim = c(7.815, 20), geom = "area", fill = "darkred") +
  geom_vline(xintercept = 7.815, colour = "darkred") + 
  geom_text(aes(x = 7.4, y = .15, label = "Critical-value = 7.815"), 
            angle = 90, color = "darkred") +
  geom_text(aes(x = 8.8, y = .15, label = "Calculated-value = 8.44"), 
            angle = 90) +
  geom_vline(xintercept = 8.44) + 
  theme_classic() +
 # scale_color_viridis(discrete = TRUE) +
  ylab(element_blank()) +
  xlab(label = "Value of the goodness-of-fit statistic")
```
